Extending K-Means Clustering with Ptolemy’s Inequality \thanks{This work is partly funded by DFG grant No. 454630593: EPIX – Efficient Ptolemaic Indexing.}

### 2025-01-14 paper/content/5_results.tex

To provide a fair comparison of the algorithms,
we measure the number of distance calculations instead of wall-clock time.
Distance calculations account for the majority of computing time in k-Means algorithms and do not depend on the implementation of the algorithms to machine code.
In comparison, execution time measurements are contingent on a number of interacting factors that render them difficult to reproduce.
These include the programming language and version, the amount and distribution of time spend on optimizing each algorithm's implementation, and the hardware architecture used to conduct the measurements.
Thus, it takes considerable effort and care to make time measurements objective, reproducible, and comparable.
Consequently, this short paper focuses on the theoretical, implementation-independent aspects of the discussed algorithms to determine whether a more extensive study is justified.

@@

\textbf{Number of Clusters.}
The full Ptolemaic algorithm consistently demonstrates significant improvements over Elkan’s algorithm as the number of clusters $k$ increases.
For instance, in the Gaussian low-dimensional dataset, the Ptolemaic algorithm achieves larger than 2-fold speedups for $k = 20$ and $k = 100$, illustrating its ability to significantly reduce distance calculations in complex clustering tasks.
% surpasses 3-fold improvements for
% Similar trends are observed in the Random low-dimensional dataset.
For smaller cluster counts ($k = 3$), the advantage is less pronounced, with speedups around 1.1 in the Iris and Wine datasets.
This suggests that Ptolemy's inequality offers significant improvements for larger values of $k$.
However, lower-complexity clustering tasks may not benefit to the same extent,
given the additional overhead imposed by the Ptolemaic bounds, which are computationally more complex to evaluate compared to the triangle bounds.
The Ptolemaic Upper Bound variant performs comparably to the full Ptolemaic algorithm, especially for larger $k$ values.
However, the Lower Bound variant underperforms in some cases, particularly with smaller cluster counts or in higher-dimensional datasets.

\textbf{Dimensionality.} The full Ptolemaic algorithm performs well across varying dataset dimensionality. In low-dimensional datasets (10 dimensions), it achieves impressive speedups, particularly at higher cluster counts.
As dimensionality increases, performance gains persist but become slightly less pronounced.
Nonetheless, even in higher-dimensional datasets, the full Ptolemaic algorithm remains effective, consistently reducing distance evaluations.
However, the Ptolemaic lower bound variant often struggles in these higher-dimensional cases, further emphasizing the robustness of the full Ptolemaic approach.

>>>>

To provide a fair comparison of the algorithms,
we measure the number of distance calculations instead of wall-clock time.
Distance calculations account for the majority of computing time in k-Means algorithms and do not depend on the specific machine-code implementation.
In comparison, execution time measurements are contingent on a number of interacting factors that render them difficult to reproduce.
These include the programming language and version, the amount and distribution of time spend on optimizing each algorithm's implementation, and the hardware used to conduct such measurements.
Thus, it takes considerable effort and care to make time measurements objective, reproducible, and comparable.
Consequently, this short paper focuses only on the theoretical, implementation-independent aspects of the discussed algorithms to determine whether a more extensive study is justified.

@@

\textbf{Number of Clusters.}
The full Ptolemaic algorithm consistently demonstrates significant improvements over Elkan’s algorithm as the number of clusters $k$ increases.
For instance, in the Gaussian low-dimensional dataset, the Ptolemaic algorithm achieves at least a 2-fold speedup for $k = 20$ and $k = 100$, illustrating its ability to significantly reduce distance calculations in complex clustering tasks.
% surpasses 3-fold improvements for
% Similar trends are observed in the Random low-dimensional dataset.
For smaller cluster counts ($k = 3$), the advantage is less pronounced: We see a significant difference in performance only in the low-dimensional Guassian dataset, while all other datasets show no significant increases or decreases.
This suggests that Ptolemy's inequality offers significant improvements only for larger values of $k$.
Lower-complexity clustering tasks might not benefit to the same extent,
given the additional overhead imposed by the Ptolemaic bounds, which are computationally more complex to evaluate compared to the triangle bounds.
The Ptolemaic Upper Bound variant performs roughly similar to the full Ptolemaic algorithm.
In comparison, the Lower Bound variant has comparable or worse performance to Elkan.

\textbf{Dimensionality.} The full Ptolemaic algorithm performs reasonably well across varying dataset dimensionality. In low-dimensional datasets (10 dimensions), it achieves impressive speedups, particularly at higher cluster counts.
As dimensionality increases, performance gains persist but become slightly less pronounced.
Nonetheless, even in higher-dimensional datasets, the full Ptolemaic algorithm remains effective, consistently reducing distance evaluations.
However, the Ptolemaic lower bound variant often struggles in these higher-dimensional cases, further emphasizing the robustness of the full Ptolemaic approach.

### 2025-01-14 paper/content/6_conclusion.tex

We see two primary avenues for future work:
Firstly, more extensive experiments are required to verify our findings.
Specifically, execution times on a wider range of datasets must be measured to determine under which conditions the extension leads to real-world runtime savings.
Secondly, exploring adaptations of this approach to other k-means algorithms and optimizations could prove valuable.
In particular for algorithms which rely on the triangle inequality to prune distance calculations, Ptolemy's inequality might be leveraged to achieve better performance, making these clustering algorithms more practical for a broader range of applications.

>>>>

We see two primary avenues for future work:
Firstly, more extensive experiments are required to verify our findings.
Specifically, execution times---instead of distance calculation counts---on a wider range of datasets must be measured to determine under which conditions the extension leads to real-world runtime savings.
Secondly, exploring adaptations of this approach to other k-means algorithms and optimizations could prove valuable.
In particular for algorithms which rely on the triangle inequality to prune distance calculations, Ptolemy's inequality might be leveraged to achieve better performance, making these clustering algorithms more practical for a broader range of applications.
